---
title: "Mapping-Census-Data"
author: "aaron mamula"
date: "8/4/2020"
output: html_document
---

# {.tabset .tabset-fade .tabset-pills}

## Purpose

Here's what's happening in this module: 

Working with the Census Bureau ACS data I'm going to create an informative map using the spatial data package *sf*. The exercise will proceed more-or-less like this:

* get the shapefile for CDP boundaries from the Census TIGER Data Mart
* join the boundry shapefile with the data series 
* create a "heatmap" of educational attainment by county

## Data and Packages {.tabset}

### Packages

```{r warning=F}
library(sf)
library(dplyr)
library(data.table)
library(ggplot2)
library(RJSONIO)
library(viridis)
library(here)
```

### Data

[For this application I used the 2016 California county boundaries GIS files from the Census Bureau's TIGER Data Mart](https://catalog.data.gov/dataset/tiger-line-shapefile-2016-state-california-current-county-subdivision-state-based). You can replace this with a more recent GIS file if you like. I chose this one because GitHub places a limit on the size of files that can be uploaded to a project. I wanted you all to have access to the data files within GitHub so I looked for a set of GIS files that could be placed within the GitHub repository without exceeding the space constraints. 

```{r}

county.lines <- st_read(here('data/CA_Counties/CA_Counties_TIGER2016.shp'))
str(county.lines)
```

The thing to note from above is that we have an data frame-looking thing that is object class *sf* with a variety of spatial data type identifiers. The [Simple Features Package](https://r-spatial.github.io/sf/articles/sf1.html) really simplifies working with spatial data because it organizes things in a familiar data-frame looking way.

## Skills

* data import with shapefiles
* working the the [Simple Features Package](https://r-spatial.github.io/sf/articles/sf1.html)
* data manipulation with dplyr
* working with APIs and JSON data
* plotting with ggplot

## A Test Map

As a first step, plot the county boundaries just to make sure everything looks ok. 

We can display these county boundaries on a map using the *ggplot()* method. 

```{r}
library(ggthemes)
ca.counties <- county.lines %>% filter(STATEFP=='06')
ggplot(ca.counties) + geom_sf() + theme_economist_white()
```

## A Real Map

In the last section I made a base map. This was a map of California county boundaries. Here, I'm going to draw on the "working with APIs" section of the Data Import Skill and pull some data from the 2018 ACS to use with the basemap.

### Get the Educational Attainment Series and Join with Spatial Boundaries

This section draws on the earlier less on using APIs to extract data. Here I'm using the [Census Bureau's ACS API](https://www.census.gov/data/developers/data-sets.html). Having worked with these data for some time, I have some prior experience with the data series codes. Here I'm going to extract the data series "% of population with a Bachelors Degree" at the county level for the state of California.


```{r}
data.series.name <- 'DP02_0064PE'

# my API key
key <- "key=f5a32f694a14b28acf7301f4972eaab8551eafda"
# the base url for Detailed Tables for the 2018 ACS 5-year
base <- "https://api.census.gov/data/2018/acs/acs5/profile?get=NAME,"
# geo parameters
geo <-  "&for=county:*&in=state:06&"

# paste these all together to form the API call for the poverty data series

  endpoint <- paste(base,data.series.name,geo,key,sep="")
  
  # get json data
  data <- fromJSON(endpoint)

  # parse json object and coerce to data frame
  edu <- tbl_df(data.frame(rbindlist(lapply(data,function(x){
    x<-unlist(x)
    return(data.frame(name=x[1],value=x[2],state=x[3],data_series=data.series.name))
  })
  ))) %>% filter(row_number() > 1)

edu$value <- as.numeric(as.character(edu$value))    
```

Next, I need to join these data on educational attainment with the spatial data on county boundaries. One challenge, illustrated below, is that the naming convention is a little different. 

County names from the Census ACS data:
```{r}
head(unique(edu$name))
```

County names from the Census boundaries shapefile:
```{r}
head(unique(county.lines$NAME[which(county.lines$STATEFP=='06')]))
```

It looks to me like if I just drop the "County, California" part from the ACS name identifiers than the two will be conformable. Let's try:

```{r}
# I'm changing the column label for the 'name' field in order to be conformable with the county lines
# data. This isn't strictly necessary for a join operation but makes things easier.
edu$NAME <- trimws(gsub("County, California","",edu$name))
head(unique(edu$NAME))
edu <- edu %>% select(-name)
```

Try joining and see what doesn't work
```{r}

ca.counties <- county.lines %>% filter(STATEFP=='06') %>% left_join(edu,by=c('NAME'))
ca.counties <- st_as_sf(ca.counties)  
```


### Create the Map

```{r}
ggplot(ca.counties) + geom_sf(aes(fill=value))  + 
 labs(fill="% of population\nwith college degree") +
    scale_fill_viridis() +
  ggtitle("Educational Attainment by County for California, 2018") +
  theme_classic()

```
